---
title: "Player and comparison-specific effects, and ties, in Bradley-Terry models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with speedyBBT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning = FALSE}
library(speedyBBT)
library(coda)
```

## Bradley-Terry model with player and comparison specific effects

In some studies, there may be specific covariates associated with each player or comparison. We model the probability that player $i$ beat player $j$ in comparison $k$ through

$$ logit(π_{ijk}) = (x_i − x_j)^Tβ + κa_k. $$
To account for player specific effects in this comparisons, player $i$ is associated with a set of $P$ explanatory variables $x_i = {x_{i1}, \ldots, x_{iP} }, which can be combined with the parameter
vector $\beta = {β1, \ldots, βP } $through a linear predictor to give the quality for player $i$, $λ_i = x^Ti\beta$.
The variable $a_k \varepsilon R$ describes the comparison specific effect for comparison $k$. One common
comparison specific effect is a home/away effect, for example in football leagues where the
home team has ad advantage. To allow for a home/away effect, we let $a_k = 1$ if player $i$ has
the advantage in comparison $k$, $a_k = −1$ if player $$j$ has the advantage, and $a_k = 0$ if neither players have an advantage.
As this model allows for comparison specific effects, we reformulate the likelihood function for the standard model to consider the likelihood contribution for each of the $K$ comparisons.
The likelihood function is therefore given by

$$
\pi(β, α2, κ | y, a1, . . . , ak) =
\prod_{k=1}^K \pi_{ijk}^{y_{ijk}} (1 - \pi_{ijk})^{1 - y_{ijk}}
$$

The prior distributions for this model are similar to the standard model and given by

\begin{eqnarray}
α2 ∼ inv − Γ(χ, ψ)
β ∼ MVN(μ, α2Σ)
κ ∼ N(0, τ 2)
z ∼ PG(1, 0)
\end{eqnarray}

## 2019 Wimbledon Championship

We demonstrate the `speedyBBT` package on data from the 2019 Wimbledon Men’s Singles Championship. This consists of 128 players playing 127 matches. We include two covariates for each player, their rank in the ATP league table and the number of points received so far in the ATP 2019 tour prior to Wimbledon starting. The quality parameter for player $i$ is
modelled by $$\lambda_i = \beta_1x_{rank,i} + \beta_2x_{points,i}$$ We include a comparison specific effect by setting
$$a_k = \begin{cases}
0 & \text{if match } k \text{ took place in week 1} \\
1 & \text{if match } k \text{ took place in week 2}
\end{cases}$$

This is to allow for player fatigue and wear of the grass courts between weeks. We fit the model by calling

```{r wimbledon}
data(wimbledon, package = "speedyBBT")

# Fit BT model with covariates and comparison-specific effect
wimbledonModel <- BBTm(
  outcome = wimbledon$matches$outcome,
  item1 = wimbledon$matches$winner,
  item2 = wimbledon$matches$loser,
  advantage = wimbledon$matches$secondWeek,
  formula = ~ rank + points,
  data = wimbledon$players,
  n.iter = 4000,
  burn.in = 100
)
```

Using a burn-in period of 100 iterations, the posterior distributions for the model parameters are given by:

```{r wimbledon_results}

# Posterior histograms (burn-in = 100)
hist(
  parameter(wimbledonModel, "kappa"),
  main = "",
  xlab = expression(kappa),
  freq = FALSE,
  breaks = 14
)
```

The posterior median for the effect of a match being in the second week is around 6. 

```{r wimbledon_results2}
hist(
  parameter(wimbledonModel, "beta", indices_to_extract = 1),
  main = "",
  xlab = expression(beta[1]),
  freq = FALSE
)
hist(
  parameter(wimbledonModel, "beta", indices_to_extract = 2),
  main = "",
  xlab = expression(beta[2]),
  freq = FALSE
)
hist(
  0.5 * log(parameter(wimbledonModel, "alpha.sq")),
  main = "",
  xlab = expression(log(alpha)),
  freq = FALSE
)
```


The function also outputs samples for the player quality parameters λ and we can use these to rank the players. The top 5 players in this tournament, with Djokovic being estimated as the best, is computed by:

```{r best_players}

par(mfrow = c(2, 2))
for (i in seq_len(length(wimbledon$players$name))) {
  hist(
    parameter(wimbledonModel, "lambda", indices_to_extract = i),
    main = "",
    xlab = paste("lambda[", wimbledon$players$name[i], "]", sep = ""),
    freq = FALSE
  )

  plot(
    parameter(wimbledonModel, "lambda", indices_to_extract = i),
    type = 'l',
    xlab = "Iteration",
    ylab = paste("lambda[", wimbledon$players$name[i], "]", sep = "")
  )
}


# Top 5 players by posterior mean quality
lambda.mean <- colMeans(parameter(wimbledonModel, "lambda", indices_to_extract = 1:5))
wimbledonResults <- data.frame(
  "Name" = tail(wimbledon$players$name[order(lambda.mean)]),
  "Posterior mean quality" = tail(sort(lambda.mean))
)

# Not in main branch yet
#plot_qualities(wimbledonModel, wimbledon$players$name)
```

## Bradley–Terry with Ties
Many comparative judgement studies feature tied comparisons. This could be a sporting league where matches result in ties, or social studies where judges are say the players have approximately equal qualities. We follow Caron and Doucet (2012) and introduce a threshold
parameter $θ ≥ 0$, the probability of player $i$ beating $j$ is specified through
$$logit(π_{ij}) = λ_i − λ_j + θ$$
Small values of $θ$ means that there is a high threshold in the relative difference resulting in a tie. This yields the probabilities of a win and a tie as:

\begin{eqnarray}
π_{ij} &=& \frac{e^{\lambda_i}}{e^{\lambda_i} + e^{\lambda_j} + e^{\theta}} \\
π_{i=j} &=& \frac{(e^{2\theta} - 1)e^{\lambda_i + \lambda_j}}{(e^{\lambda_i} + e^{\lambda_j} + e^{\theta})(e^{\lambda_i + \theta} + e^{\lambda_j})}
\end{eqnarray}

As in the other routines, it places a multivariate normal prior distribution on the player quality
parameters $λ$ and an inverse-gamma prior distribution on the variance hyperparameter of the
multivariate normal distribution. We place an exponential distribution on the threshold
parameter $θ$.
$$α2 ∼ inv − Γ(χ, ψ)$$
$$λ ∼ MVN(μ, Σ)$$
$$z ∼ PG(1, 0)$$
$$θ ∼ Exp(τ )$$

In the MCMC algorithm, all parameters except for $θ$ are updated by drawing from their full conditional distribution. The the threshold parameter $θ$ is updated using a Metropolis-Hastings random walk algorithm, where new values of $θ$ are proposed from a normal distribution with mean the current value of theta and variance $ν^2$. A full description of the underlying
statistical framework can be found in Seymour and Hernandez (2024).

```{r tied_model}

```
## References

Caron F, Doucet A (2012). “Efficient Bayesian Inference for Generalized Bradley–Terry Models.” Journal of Computational and Graphical Statistics, 21(1), 174–196. doi:10.1080/10618600.2012.638220.

Seymour RG, Hernandez F (2024). “Scalable Bayesian Inference for  Bradley–Terry Models with Ties: An Application to Honour Based Abuse.” 2405.13399, URL https://arxiv.org/abs/2405.13399. 