---
title: "Player and comparison-specific effects, and ties, in Bradley-Terry models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with speedyBBT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning = FALSE}
library(speedyBBT)
library(coda)
```

## Bradley-Terry model with player and comparison specific effects

In some studies, there may be specific covariates associated with each player or comparison. We model the probability that player $i$ beat player $j$ in comparison $k$ through

$$ logit(π_{ijk}) = (x_i − x_j)^Tβ + κa_k. $$
To account for player specific effects in this comparisons, player $i$ is associated with a set of $P$ explanatory variables $x_i = {x_{i1}, \ldots, x_{iP} }, which can be combined with the parameter
vector $\beta = {β1, \ldots, βP } $through a linear predictor to give the quality for player $i$, $λ_i = x^Ti\beta$.
The variable $a_k \varepsilon R$ describes the comparison specific effect for comparison $k$. One common
comparison specific effect is a home/away effect, for example in football leagues where the
home team has ad advantage. To allow for a home/away effect, we let $a_k = 1$ if player $i$ has
the advantage in comparison $k$, $a_k = −1$ if player $$j$ has the advantage, and $a_k = 0$ if neither players have an advantage.
As this model allows for comparison specific effects, we reformulate the likelihood function for the standard model to consider the likelihood contribution for each of the $K$ comparisons.
The likelihood function is therefore given by

\eqarray{
    \pi(β, α2, κ | y, a1, . . . , ak) =
\prod_{k=1}^K \pi_{ijk}^{y_{ijk}} (1 - \pi_{ijk})^{1 - y_{ijk}}
}
The prior distributions for this model are similar to the standard model and given by
\eqnarray{
α2 ∼ inv − Γ(χ, ψ)
β ∼ MVN(μ, α2Σ)
κ ∼ N(0, τ 2)
z ∼ PG(1, 0)
}

## 2019 Wimbledon Championship
We demonstrate the speedyBBT package on data from the 2019 Wimbledon Men’s Singles
Championship. This consists of 128 players playing 127 matches. We include two covariates
for each player, their rank in the ATP league table and the number of points received so far
in the ATP 2019 tour prior to Wimbledon starting. The quality parameter for player i is
modelled by
λi = β1xrank,i + β2xpoints,i
We include a comparison specific effect by setting
ak =
(
0 if match k took place in week 1
1 if match k took place in week 2
. This is to allow for player fatigue and wear of the grass courts between weeks. We fit the
model by calling
```{r wimbledon}
data(wimbledon, package = speedyBBT)

# Fit BT model with covariates and comparison-specific effect
wimbledonModel <- BBTm(
  outcome = wimbledon$matches$outcome,
  item1 = wimbledon$matches$winner,
  item2 = wimbledon$matches$loser,
  advantage = wimbledon$matches$secondWeek,
  formula = ~ rank + points,
  data = wimbledon$players,
  n.iter = 4000,
  burn.in = 100
)
```

Using a burn-in period of 100 iterations, the posterior distributions for the model parameters
are given by:

```{r wimbledon_results}

# Posterior histograms (burn-in = 100)
hist(
  parameter(wimbledonModel, "kappa"),
  main = "",
  xlab = expression(kappa),
  freq = FALSE,
  breaks = 14
)
```

The posterior median for the effect of a match being in the second week is around 6. 
```{r wimbledon_results2}
hist(
  parameter(wimbledonModel, "beta", indices_to_extract = 1),
  main = "",
  xlab = expression(beta[1]),
  freq = FALSE
)
hist(
  parameter(wimbledonModel, "beta", indices_to_extract = 2),
  main = "",
  xlab = expression(beta[2]),
  freq = FALSE
)
hist(
  0.5 * log(parameter(wimbledonModel, "alpha.sq")),
  main = "",
  xlab = expression(log(alpha)),
  freq = FALSE
)
```


The function also outputs samples for the player quality parameters λ and we can use these to rank the players. The top 5 players in this tournament, with Djokovic being estimated as the best, is computed by:
```{r best_players}

par(mfrow = c(2, 2))
for (i in seq_len(length(wimbledon$players$name))) {
  hist(
    parameter(wimbledonModel, "lambda", indices_to_extract = i),
    main = "",
    xlab = paste("lambda[", wimbledon$players$name[i], "]", sep = ""),
    freq = FALSE
  )

  plot(
    parameter(wimbledonModel, "lambda", indices_to_extract = i),
    type = 'l',
    xlab = "Iteration",
    ylab = paste("lambda[", wimbledon$players$name[i], "]", sep = "")
  )
}


# Top 5 players by posterior mean quality
lambda.mean <- colMeans(parameter(wimbledonModel, "lambda", indices_to_extract = 1:5))
wimbledonResults <- data.frame(
  "Name" = tail(wimbledon$players$name[order(lambda.mean)]),
  "Posterior mean quality" = tail(sort(lambda.mean))
)

# Not in main branch yet
#plot_qualities(wimbledonModel, wimbledon$players$name)
```

```{r tied_model, eval = FALSE}

###############################################################################
# Section 4.1: Deprivation in Dar es Salaam — Bradley-Terry with Ties
###############################################################################

# Inspect comparison data
head(darEsSalaam$comparisons)

# Construct prior covariance matrix
prior.var <- expm::expm(darEsSalaam$adjacencyMatrix)
prior.var <- diag(diag(prior.var)^-0.5) %*%
  prior.var %*%
  diag(diag(prior.var)^-0.5)

# Fit tied BT model
darTiedModel <- BBTm.ties(
  outcome = darEsSalaam$comparisons$outcome,
  item1 = darEsSalaam$comparisons$subward1,
  item2 = darEsSalaam$comparisons$subward2,
  item.prior.var = prior.var,
  hyperparameter = TRUE,
  rw.sd = 0.005,
  n.iter = 5000,
  burn.in = 100
)

# Trace plot and histogram for threshold parameter theta
par(mfrow = c(1, 2))
plot(
  parameter(darTiedModel, "theta"),
  type = 'l',
  xlab = "Iteration",
  ylab = expression(theta)
)
hist(
  parameter(darTiedModel, "theta"),
  main = "",
  xlab = expression(theta),
  freq = FALSE
)

# Posterior mean and 95% credible interval for theta
mean(parameter(darTiedModel, "theta"))
quantile(parameter(darTiedModel, "theta"), c(0.025, 0.975))
```
